A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

### 2020-05-09 Manuscript/Manuscript-6-DCA.tex
vsn 6

@@

Investigating macroevolutionary patterns of trait variation requires a
phylogenetic perspective, because the shared ancestry among species
generates statistical non-independence (Felsenstein 1985; Harvey and
Pagel 1991). Accounting for this evolutionary non-independence is the
purview of \emph{phylogenetic comparative methods} (PCMs); a suite of
analytical tools that condition the data on the phylogeny through the
course of statisical evaluations of phenotypic trends (e.g., Grafen
1989; Garland and Ives 2000; Rohlf 2001; Butler and King 2004). The past
several decades have witnessed a rapid expansion in the development of
PCMs to address an ever-growing set of macroevolutionary hypotheses
(Martins and Hansen 1997; O'Meara et al. 2006; Revell and Harmon 2008;
Beaulieu et al. 2012; Adams 2014b,a; Adams and Collyer 2018). These
methods are predicated on the notion that phylogenetic signal -- the
tendancy for closely related species to display similar trait values --
is present in cross-species datasets (Felsenstein 1985; Pagel 1999;
Blomberg et al. 2003). Indeed, under numerous evolutionary models,
phylogenetic signal is to be expected, as stochastic character change
along the hierarchical structure of the tree of life generates trait
covaration among related taxa (see Felsenstein 1985; Blomberg et al.
2003; Revell et al. 2008). \hfill\break

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(Kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Munkemuller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). Here, maximum likelihood is used to fit the data to the
phylogeny under a Brownian motion model of evolution. A parameter
(\(\lambda\)) is included, which transforms the lengths of the internal
branches of the phylogeny to improve the fit (Pagel 1999; Freckleton et
al. 2002). Pagel's \(\lambda\) ranges from \(0\to1\), with larger values
signifying a greater dependence of observed trait variation on the
phylogeny. Pagel's \(\lambda\) also has the appeal that it may be
included in phylogenetic regression (PGLS) to account for the degree of
phylogenetic signal in comparative analyses (see Freckleton et al.
2002). \hfill\break

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. For instance, of the 204
papers published in 2019 that estimated and reported Pagel's \(\lambda\)
(found in Google.scholar), 40\% interpreted the strength of phylogenetic
signal for at least one phenotypic trait. Additionally, because nearly
half of the 1,572 \(\lambda\) values reported were near the limits of
the parameter (Figure 1), this percentage is even higher, as biological
interpretation of phylogenetic signal at the limits of \(\lambda\) are
known. \hfill\break

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also assumes that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we still
lack a general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure the magnitude of a statistical effect in data, represented on a
common scale (Glass 1976; Cohen 1988). Effect sizes have widespread use
in many areas of the quantiative sciences, as they represent measures
that may be readily summarized across datasets as in meta-analysis
(Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster 1995), or
compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and test statistics are effect
sizes, and thus many summary measures must first be converted to
standardized units (i.e., an effect size) for meaningful comparison (see
Rosenthal 1994). As a consequence, it follows that only if \(\lambda\)
is a statistical effect size can comparisons of estimates across
datasets be interpretable. For the case of \(\lambda\), this has not yet
been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable estimate of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(Kappa\). Through simulations
across a wide range of conditions, we find that the precision of effect
sizes based on \(\lambda\) (\(Z_{\lambda}\)) are less reliable than that
those based on \(Kappa\) (\(Z_K\)), implying that \(Z_K\) is a more
robust effect size measure. We also propose a two-sample test statistic
that may be used to compare the strength of phylogenetic signal among
datasets, and provide an empirical example to demonstrate its use. We
conclude that estimates of phylogenetic signal using Pagel's \(\lambda\)
are often inaccurate, and thus interpreting strength of phylogenetic
signal in phenotypic datasets based on this measure is compromised. By
contrast, effect sizes obtained from \(Kappa\) hold promise for
characterizing phylogenetic signal, and for comparing the strength of
phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal corresponding reflecting Brownian motion (when
\(\lambda_{in} =1\)). For each dataset we then estimated phylogenetic
signal (\(\lambda_{est}\)), and calculated the precision of \(\lambda\)
using the variance (\(\sigma^2_\lambda\)) across datasets at each input
level of phylogenetic signal and level of species richness. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each phylogeny under a Brownian motion model of
evolution (for PGLS regression). For phylogenetic ANOVA, random groups
(\(X\)) were obtained by simulating a discrete (binary) character on
each phylogeny. Next, the dependent variable was simulated in such a
manner as to contain a known relationship with \(X\) plus random error
containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). Here, the association between \(Y\) and \(X\)
was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). The fit of the
phylogenetic regression was estimated using maximum likelihood, and
parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained. Precision estimates (\(\sigma^2_\lambda\)) at each input level
of phylogenetic signal and level of species richness were then observed.
\hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
obtained in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these empirical estimates were explicitly interpreted in terms
of the strength of phylogenetic signal that they represented (i.e.,
weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression, the results of our simulations were quite similar.
Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3), and the spread of \(\lambda_{est}\) was
similar to that observed when \(\lambda\) was estimated for only the
dependent variable, as in Figure 2. Taken together, these findings
reveal that \(\lambda_{est}\) does not precisely characterize observed
levels of phylogenetic signal in phenotypic datasets, and that
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) may be highly inaccurate. \hfill\break

@@

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (Glass 1976; Cohen 1988; Rosenthal 1994).
\(Z_{\theta}\) expresses the magnitude of the effect in \(\theta_{obs}\)
by transforming the original test statistic to its standard normal
deviate (Glass 1976; Kelley and Preacher 2012). Typically,
\(\theta_{obs}\) and \(\sigma_\theta\) are estimated from the data,
while \(E(\theta)\) is obtained from the distribution of \(\theta\)
derived from parametric theory. However, recent advances in resampling
theory (Collyer et al. 2015; Adams and Collyer 2016, 2019a) have shown
that \(E(\theta)\) and \(\sigma_\theta\) may also be obtained from an
empirical sampling distribution of \(\theta\) obtained from permutation
procedures. \hfill\break

Adams and Collyer (2019b) suggested that the strength of phylogenetic
signal could be represented as an effect size, based on the \(Kappa\)
statistic and its empirical sampling distribution from permutation. Here
we formalize that suggestion, resulting in an effect size of:

\begin{align}
    Z_{K}=\frac{K_{obs}-\hat\mu_K}{\hat\sigma_K}
\end{align}

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \(Kappa\) obtained via permutation.
Similarly, an effect size based on \(\lambda\) could be envisioned as:

@@

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood. Note that under the null
hypothesis, no phylogenetic signal is expected (Freckleton et al. 2002),
and thus \(E(\lambda)=0\) under this condition. \hfill\break

>>>>

Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. In this study, we evaluated the
efficacy of one commonly used parameter (Pagel's \(\lambda\)) to
estimate the strength of phylogenetic signal in phenotypic traits, and
evaluate the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal. We find that the precision of \(\lambda\)
in estimating actual levels of phylogenetic signal is often inaccurate,
and that biological interpretations of the strength of phylogenetic
signal based on \(\lambda\) are therefore compromised. We then propose a
standardized effect size based on \(\kappa\) (\(Z_\kappa\)), which
measures the strength of phylogenetic signal, and places it on a common
scale for statistical comparison. Tests based on \(Z_\kappa\) provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scale.

@@

Investigating macroevolutionary patterns of trait variation requires a
phylogenetic perspective, because the shared ancestry among species
violates an assumption of independence among trait values that is common
for statistical tests (Felsenstein 1985; Harvey and Pagel 1991).
Accounting for this evolutionary non-independence is the purview of
\emph{phylogenetic comparative methods} (PCMs): a suite of analytical
tools that condition trends in the data on the phylogenetic relatedness
of observations (e.g., Grafen 1989; Garland and Ives 2000; Rohlf 2001;
Butler and King 2004). The past several decades have witnessed a rapid
expansion in the development of PCMs to address an ever-growing set of
macroevolutionary hypotheses (Martins and Hansen 1997; O'Meara et al.
2006; Revell and Harmon 2008; Beaulieu et al. 2012; Adams 2014b,a; Adams
and Collyer 2018). These methods are predicated on the notion that
phylogenetic signal -- the tendancy for closely related species to
display similar trait values -- is present in cross-species datasets
(Felsenstein 1985; Pagel 1999; Blomberg et al. 2003). Indeed, under
numerous evolutionary models, phylogenetic signal is to be expected, as
stochastic character change along the hierarchical structure of the tree
of life generates trait covaration among related taxa (see Felsenstein
1985; Blomberg et al. 2003; Revell et al. 2008). \hfill\break

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Munkemuller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The (\(\lambda\)) parameter transforms the lengths of the
internal branches of the phylogeny to improve a fit of data to the
phylogeny via maximum likelihood (Pagel 1999; Freckleton et al. 2002).
Pagel's \(\lambda\) ranges from \(0\to1\), with larger values signifying
a greater dependence of observed trait variation on the phylogeny.
Pagel's \(\lambda\) also has the appeal that it may be included in
phylogenetic generalized least-squares regression (PGLS) to account for
the degree of phylogenetic signal in comparative analyses (see
Freckleton et al. 2002). \hfill\break

In addition to functioning as a parameter that is tuned for appropriate
analysis, \(\lambda\) can function as a descriptive statistic to
describe the relative strength of phylogenetic signal in phenotypic
traits, to determine the extent to which shared evolutionary history has
influenced trait covariation among taxa. The appeal of \(\lambda\) as a
descriptive statistic for evolutionary biologists is a basis for
interpreting ``weak'' versus ``strong'' phylogenetic signal; i.e., small
versus large values of \(\lambda\), respectively, in a comparative sense
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Indeed, statements regarding the strength of phylogenetic signal based
on \(\lambda\) are rather common in the evolutionary literature. For
instance, of the 204 papers published in 2019 that estimated and
reported Pagel's \(\lambda\) (found from a literature survey we
conducted in Google.scholar), 40\% interpreted the strength of
phylogenetic signal for at least one phenotypic trait. Further, because
nearly half of the 1,572 \(\lambda\) values reported were near 0 or 1
(Figure 1) where the biological interpretation of \(\lambda\) is known,
this percentage is even higher. \hfill\break

@@

Various other approaches use \(\lambda\) as a parameter that can be
varied for inferences akin to sensitivity analysis. For instance, some
have performed likelihood ratio tests that compare observed model fits
to those obtained when \(\lambda=0\) or \(\lambda=1\) (Freckleton et al.
2002; Cooper et al. 2010; Bose et al. 2019) or evaluated whether
observed \(\lambda\) differs from an expected \(\lambda\), based on
confidence intervals generated for the expected value (Vandelook et al.
2019). Qualitative comparisons of \(\lambda\) estimates have also been
performed for multiple traits on the same phylogenetic tree to infer
whether the strength of phylogenetic signal is greater in one trait as
compared to another (e.g., Liu et al. 2019; Bai et al. 2019).
\hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also implies that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we lack a
general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and descriptive statistics are
effect sizes, and thus many summary measures must first be converted to
statistics with standardized units (i.e., conversion to an effect size)
for meaningful comparison (see Rosenthal 1994). As a consequence, it
follows that only if \(\lambda\) is a statistical effect size can
comparisons of estimates across datasets be interpretable. For the case
of \(\lambda\), this has not yet been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable indicator of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(\kappa\). Through simulations we
find that the precision of effect sizes based on \(\lambda\)
(\(Z_{\lambda}\)) are less reliable than that those based on \(\kappa\)
(\(Z_\kappa\)), implying that \(Z_\kappa\) is a more robust effect size
measure. We also propose a two-sample test statistic that may be used to
compare the strength of phylogenetic signal among datasets, and provide
an empirical example to demonstrate its use. We conclude that estimates
of phylogenetic signal using Pagel's \(\lambda\) are often inaccurate,
and thus interpreting strength of phylogenetic signal in phenotypic
datasets based on this measure is compromised. By contrast, effect sizes
obtained from \(\kappa\) hold promise for characterizing phylogenetic
signal, and for comparing the strength of phylogenetic signal across
datasets.

@@

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal reflecting Brownian motion (when
\(\lambda_{in} =1\)). For each dataset we then estimated phylogenetic
signal (\(\lambda_{est}\)), and calculated the variance of \(\lambda\)
(\(\sigma^2_\lambda\)) across datasets at each input level of
phylogenetic signal and level of species richness as an estimate of
precision. We verified that the variance of traits simulated had no
effect on phylogenetic signal estimation. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each rescaled phylogeny under a Brownian motion
model of evolution (for PGLS regression). For phylogenetic ANOVA, random
groups (\(X\)) were obtained by simulating a discrete (binary, 0 or 1)
character on each phylogeny. Next, the dependent variable was simulated
in such a manner as to contain a known relationship with \(X\) plus
random error containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). The association between \(Y\) and \(X\) was
modeled using a range of values: \(\beta=(0.0,0.25, 0.5, 0.75,1.0)\),
and the residual error (\(\epsilon\)) was modeled to contain
phylogenetic signal simulated under a Brownian motion model of evolution
on each rescaled phylogeny:
\(\epsilon=\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\): (see Revell
2010 for a similar simulation design). The fit of the phylogenetic
regression was estimated using maximum likelihood, and parameter
estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were obtained. We then
calculated precision estimates (\(\sigma^2_\lambda\)) at each input
level of phylogenetic signal and level of species richness. We verified
that the amount of residual variance simulated had no effect on
\(\sigma^2_\lambda\) but did influence the precision of coefficients
estimated from the linear model (precision increased with smaller
\(\epsilon\), as expected). \hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
published in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression and ANOVA, the results of our simulations were quite
similar. Regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3; see also Supporting Information), and the
spread of \(\lambda_{est}\) was similar to that observed when
\(\lambda\) was estimated for only the dependent variable, as in Figure
2. Taken together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (Glass 1976; Cohen 1988; Rosenthal 1994).
\(Z_{\theta}\) expresses the magnitude of the effect in \(\theta_{obs}\)
by transforming the original test statistic to its standard normal
deviate (Glass 1976; Kelley and Preacher 2012). Typically,
\(\theta_{obs}\) and \(\sigma_\theta\) are estimated from the data,
while \(E(\theta)\) is obtained from the distribution of \(\theta\)
derived from parametric theory. However, recent advances in resampling
theory (Collyer et al. 2015; Adams and Collyer 2016, 2019a; Collyer and
Adams 2018) have shown that \(E(\theta)\) and \(\sigma_\theta\) may also
be obtained from an empirical sampling distribution of \(\theta\)
obtained from permutation procedures. \hfill\break

Adams and Collyer (2019b) suggested that the strength of phylogenetic
signal could be represented as an effect size based on the \(\kappa\)
statistic and its empirical sampling distribution from permutation. Here
we formalize that suggestion, resulting in an effect size of:

\begin{align}
    Z_\kappa=\frac{log(\kappa_{obs})-\hat\mu_{log(\kappa)}}{\hat\sigma_{log(\kappa)}}
\end{align}

where \(\kappa_{obs}\) is the observed phylogenetic signal, and
\(\hat\mu_\kappa\) and \(\hat\sigma_\kappa\) are the mean and standard
deviation of the empirical sampling distribution of \(log(\kappa)\)
obtained via permutation. Note that the logarithm was used because
\(\kappa\) takes only positive values (\(0\to\infty\)) and its sampling
distribution is log-normally distributed (for a similar transformation
when calculating multivariate effect sizes see: Appendix 1 of Collyer
and Adams 2018). \hfill\break

An effect size based on \(\lambda\) could be envisioned, which is found
as:

@@

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood, as permutation approaches
have not been developed for evaluating \(\lambda\). Note also that under
the null hypothesis, no phylogenetic signal is expected (Freckleton et
al. 2002), and thus \(E(\lambda)=0\) under this condition. \hfill\break

To evaluate the utility of \(Z_\kappa\) and \(Z_\lambda\) we calculated
both effect sizes for the simulated datasets generated above, and
summarized the precision of each using its variance
(\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\), Figure 4:
additional results in the Supporting Information). Here two things are
evident. First, estimates of \(Z_\kappa\) linearly track the input
phylogenetic signal whereas estimates of \(Z_\lambda\) do not (Figure
4A, B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_\kappa\). Second, the precision of \(Z_\kappa\) is considerably more
stable as compared with \(Z_\lambda\). This may be seen by calculating
the coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_\kappa}\) and \(\sigma^2_{Z_\lambda}\)) across input
levels of phylogenetic signal. Coefficients of variation in the
precision of \(Z_\kappa\) were up to an order of magnitude smaller for
than for \(Z_\lambda\) (Figure 4C), implying that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_\kappa\). \hfill\break
